Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-08-05 paper.tex
polishing

\section{Introduction}
We present an introduction to dependent types and (Martin-L\"of) type theory.
At its core, type theory seeks to serve as a possible foundation of mathematics,
one that is based on a different logical foundation. This proof-relevant
form of mathematics is the basis in which types can be used to model logical
propositions and is the core idea of type theory. As such, not only does type
theory have a viable application in mathematics but also a growing interest
in the field of computer science. In both fields, type theory provides a richer
world of types and expressibiilty.

The study of type theory goes hand in hand with the study of dependent types,
which is why we first take the time to do familiarize ourselves with them
(Section 2). We then continue with the foundational concepts of type theory of
Per Martin-L\"of \cite{}. For this, we recommend using a proof assistant, such
as Agda. In our examples, we refer to Agda syntax rules.

@@

What follows is a brief description of each form of dependency and the
corresponding \(\lambda\)-calculus that is associated with it. With this, we
introduce the (fully-dependent) system of the \textit{Calculus of Constructions}
(\(\lambda C\)) \cite{}. Here, the word \textit{fully} signifies that \(\lambda C\)
features all four forms of dependency mentioned above.

@@

The STLC has limited expressibilty but has the benefit of removing all
\textit{infinite} calculations from the system \cite{} and is thus the
appropriate foundation for deriving a type theoretic system. This is because
the STLC lends itself quite seamlessly as component of a larger, more expressive
system. This flexibility is what we exploit to derive \(\lambda C\). To do this,
we first extend the STLC with more expressive types through simple extentions on
its grammar for simple types.

The key difference between the STLC and \(\lambda C\) are the places in which
terms and types are allowed to be. In the STLC, terms and types are entirely
seperate entities in that terms only exist on {\em term level} and types
on {\em type level}. We return to the idea of {\em levels} later. Thus, as we
mentioned earlier, this puts a constraint on the system to only allow
\(\lambda\)-abstraction and function application to happen at term level
(viz. definition of simple type). For example, the \textit{constant function}
over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC is defined:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
const : ℕ → 𝔹 → ℕ
const = λ n b → n
\end{code}
\end{minipage}
\end{center}
Given that {\tt const} is of type {\tt ℕ → 𝔹 → ℕ}, the term that inhabits this
type is {\tt λ n b → n}. In this example, we can see that terms and types are
treated differently and are only related in terms of {\em inhabitance}. This
relation is the reason for the limited expressivity of the STLC and is also the
method in which the STLC can be extended into \(\lambda C\).

@@

Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
\(\lambda2\) simply as an extension of the STLC. This system is adds
{\em polymorphic} terms to the STLC (viz. Dependency (2)). For example,
let's take the definition of {\tt const} above and change it to define
{\tt polyConst}:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
polyConst : (A B : Set) → A → B → A
polyConst A B = λ a b → a
-- Works for any type A and B
\end{code}
\end{minipage}
\end{center}

Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
are types. These two types (viz. {\tt A} and {\tt B}) are the corresponding
types of the third and fourth arguments of {\tt polyConst}. To clarify, we can
instantiate {\tt polyConst} with \(\mathbb{N}\) and \(\mathbb{B}\) resulting
in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us:
$${\tt const} = {\tt polyConst(ℕ,𝔹)}.$$
Thus, adding polymorphism to the STLC has the effect of extending the notion
of type and term. With \(\lambda2\), we are now allowed to pass types as arguments
at term level (viz. we pass the actual types \(\mathbb{N}\) and \(\mathbb{B}\)).
It is, however, necessary to constrain this expressivity with a more expressive
type (e.g. {\tt (x : Set) → ...}). Types that appear in this fashion are
called \(\Pi\)-types. Taking the grammar for simple types, we can add \(\Pi\)-types,
resulting in the grammar for types ({\tt t}) in \(\lambda2\):

@@

\(\Pi\)-types are actually significantly more expressive than the ones in
\(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}).

%% In a later system, we
%% allow \(\Pi\)-types to be parameterized over terms (e.g. {\tt ∀ (x : t) → t}).

@@

\begin{center}
\begin{minipage}{0.65\textwidth}
\begin{code}
Prim || t → t || ∀ (x : Set) → t || Set
\end{code}
\end{minipage}
\end{center}
The curious effect of doing this allows us to construct expressions of type
{\tt t → t}, where either {\tt t} can have subexpressions containing {\tt Set}.
Let's take the familiar definition of the list data structure:

@@

We discuss {\tt List} more in depth in a later section. For now, we simply ask
{\em what is the type of {\tt List}?} To answer this question, we can simply
look at the first line of the definition above (viz. {\tt List (A : Set) : Set}).
From this line, we can see that {\tt List} is a type parameterized over any type
{\tt A}. Another way to think about this statement is to think of {\tt List} as
a form of polymorphic function of type {\tt Set → Set}, as is similar with other
{\em parameterized types}. Unlike like {\tt const} or {\tt polyConst}, this
function can only be applied at type level (thus the name {\em type constructor}).
This is to insure that we cannot write type expressions that fail to instantiate
{\tt List}. For example, the type expression {\tt List → A} is incorrect since
{\tt List} cannot appear in a type expression without first being instantiated.
To correct this, we can write this type expression as: {\tt List A → A}.

@@

To review, if we compile all the previously mentioned systems above, we have
the system \(\lambda\omega\) (System F\(\omega\)). Our final addition, the
system of \(\lambda P\), allows us to construct {\em type families} which are
types themselves and are meant to directly correspond to propositions in
higher-order logic.

@@

\begin{center}
\begin{minipage}{0.7\textwidth}
\begin{code}
data _isEven : ℕ → Set where
  ZEven : 0 isEven
  NEven : ∀ n → n isEven → (Suc (Suc n)) isEven
\end{code}
\end{minipage}
\end{center}
Here, we've defined two ways in which {\tt n isEven} for all natural numbers
({\tt n}) can be proven. Returning to the idea of proof-relevant mathematics,
we do this by use of its two type constructors {\tt ZEven} and {\tt NEven},
corresponding to the two constructors of \(\mathbb{N}\). {\tt ZEven} is the
proof that 0 is an even number. {\tt NEven} then defines the property that if
a given natural number ({\tt n}) is even (viz. {\tt n isEven}), then {\tt n + 2}
is even (viz. {\tt (Suc (Suc n)) isEven}). To clarify how this type can be used,
let's prove that 4 is even. We start by providing the appropriate type definition
and also providing a name for our proof (viz. {\tt 4isEven}):
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
\end{code}
\end{minipage}
\end{center}
From the definition of {\tt isEven}, we know that to prove 4 is even, we must
know that 2 is even. Consequently, to prove 2 is even, we must know that 0 is
even. Fortunately, we have a proof that 0 is even: {\tt ZEven}. Thus, the proof
that 2 is even is {\tt NEven 0 ZEven}, and finally the proof that 4 is even is:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
4isEven = NEven 2 (NEven 0 ZEven)
\end{code}
\end{minipage}
\end{center}

@@

While it might seem intuitive and more effective to instead try to prove that 5
is {\em not} even, this, however, also doesn't work. This is due to the
treatment of negation in Intuitionistic Logic. To prove that 5 is not even is
to prove that \(\neg (5\) {\tt isEven}), which is equivalent to constructing a
function of type:
\begin{center}
\begin{minipage}{0.3\textwidth}
\begin{code}
5 isEven → ⊥
\end{code}
\end{minipage}
\end{center}
That is, given a proof that {\tt 5 isEven}, construct \(\perp\), the empty type.
This is not possible since we cannot ever construct a proof that 5 is even. We
discuss the empty type later in Section 3.

@@

\section{Martin-L\"of Type Theory}
%% Curry-Howard?
In this section, we introduce the foundational concepts of Martin-L\"of Type
Theory \cite{}. We discuss recursion, induction, and as well several primitive type
encodings (from propositional logic).

@@

\begin{minipage}{0.7\textwidth}
\begin{small}
The recursion and induction principles for some type {\tt T}.
\end{small}
\end{minipage}
\end{center}
The first argument given to these principles is sometimes called the {\em motive}.
This simply means that {\tt C} is the type from which we will return an element
from. For recursion principles, this means the given output of the function we
are defining, and for induction principles, this can be read as the proposition
we are trying to prove. The terms that follow (viz. {\tt C₀}) correspond to the
constructor(s) of the given type, which is how these principles guarantee
totality. Finally, both principles take a term of the type that are acting on
(viz. {\tt T} and {\tt (t : T)}). The only difference is that the induction
principle introduces a type family, as opposed to a non-parameterized type in the
recursion principle. We discuss this in Section 4.

@@

prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
prop₀ ¬¬¬a a = ?
\end{code}
\end{minipage}
\end{center}
Now, we solve for {\tt ?}. Looking at the type expression of {\tt prop₀},
we know we have an expression of type \(\neg\neg\neg A\) and also an
expression of type \(A\). To clarify, we can rewrite the type expression
after desugaring \(\neg A\):
$$\neg\neg\neg A\rightarrow A\rightarrow\perp$$
Now, to construct an element of type \(\neg A\) from an element of type
\(\neg\neg\neg A\), we can use the definition of \(\neg\):
$$(\neg\neg A\rightarrow\perp)\rightarrow A\rightarrow\perp$$
$$((\neg A\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
$$(((A\rightarrow\perp)\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
Thus, we know we must apply our expression \(\neg\neg\neg A\) to a function
of type \(\neg\neg A\) and so on, such that:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
prop₀ ¬¬¬a a = ¬¬¬a (λ ¬a → ¬a a)
\end{code}
\end{minipage}
\end{center}
Or equivalently:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
prop₀ ¬¬¬a a = rec⊥ ⊥ (¬¬¬a (λ ¬a → ¬a a))
\end{code}
\end{minipage}
\end{center}
Next, we introduce the type \(\top\), which contains exactly one element,
{\tt *}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data ⊤ : Set where
  * : ⊤

@@

ind⊤ : (C : ⊤ → Set) → C * → (x : ⊤) → C x
ind⊤ C c * = c
\end{code}
\end{minipage}
\end{center}
Similar to \(\perp\), \(\top\) is simple and is rather uninteresting by itself.
There are, however, some trivial yet interesting proofs that involve the two.
For the first of these, we also introduce a special kind of proof: {\it uniqueness
of identity} (UIP) \cite{}. In this proof, we set out to determine and classify
the identity of every element within a given type. For example, for \(\top\), we
should be able to prove:
$$\forall (x : \top) \rightarrow x ≡ *$$
That is, for every element of type \(\top\), that element is equivalent to {\tt *}.
This second expression, {\it x ≡} {\tt *}, is an example of an {\it identity type},
which we describe later on in this section. For now, we simply say that the
sole constructor for the identity type is {\tt (refl x)}, which is a proof that
{\it x ≡ x}.

Now, to prove the above proposition, we can use the induction principle for
\(\top\), since the proposition we are trying to prove concerns all elements
of \(\top\):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ : (x : ⊤) → x ≡ *
uip⊤ = ind⊤ (λ top → top ≡ *) (refl *)
\end{code}
\end{minipage}
\end{center}
Here, we have taken our motive to be the expression {\tt (λ top → top ≡ *)},
which essentially repeats the information already given by our type expression.
The second element, {\tt (refl *)}, is a proof that our motive holds in the case that
{\tt top} is {\tt *} and is of type {\tt * ≡ *}.

@@

ind+ : {A B : Set} (C : A + B → Set) →
       (∀ (x : A) → C (inl x)) →
       (∀ (x : B) → C (inr x)) →
       (x : A + B) → C x
ind+ C f _ (inl a) = f a
ind+ C _ f (inr b) = f b
\end{code}
\end{minipage}
\end{center}
It might come as surprise as to why the symbol for addition is generally used
for the disjunction type. This is because another name for + is a {\it union type}.
This type is parameterized over two other types, {\tt A} and {\tt B}, and
creates a union between them in that every expression of type {\tt A + B} is
either {\tt (inl a)} or {\tt (inr b)}, where {\tt a : A} and {\tt b : B}.

@@

We can now write a simple proof that shows the union of \(\top\) and \(\top\)
is equivalent to the Boolean type, that is {\tt (⊤ + ⊤) ↔ 𝔹}.
Alternatively, since \(\top\) is occasionally written as 1 and {\tt 𝔹} as 2, we
name our proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
1+1=2 : (⊤ + ⊤) ↔ 𝔹
1+1=2 = ?
\end{code}
\end{minipage}
\end{center}
To do this, we must associate every possible term of type {\tt ⊤ + ⊤} to a
given Boolean. This works out quite simply since the type {\tt ⊤ + ⊤} only has
two possible terms, {\tt (inl *)} and {\tt (inr *)}, which we can associate with
the two terms, {\tt True} and {\tt False}, in the type {\tt 𝔹} and vice-versa.
Thus, we can write the proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
1+1=2 = <> (f , g)
  where f = rec+ 𝔹 (λ inl* → True) (λ inr* → False)
        g = rec𝔹 (⊤ +' ⊤) (inl *) (inr *)
\end{code}
\end{minipage}
\end{center}
For our purposes, we use the constructor {\tt <>} for constructing a term of
the bijection type. For simplicity, we can define a bijection as a conjunction
of two logical implications:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _↔_ (A B : Set) : Set where
  <>_ : (A → B) × (B → A) → (A ↔ B)
\end{code}
\end{minipage}
\end{center}
To clarify, this form of equivalence is called ``judmental equality'' (i.e.,
equality by logical bijection) \cite{}. We go over the three different forms of
equivalence in the following section.

@@

\subsection*{Equality and Identity Types}
%% talk about definitional equality, judgmental equality, prop equality
The first form of equality is much simpler than the others. This form of
equality is called {\it definitional equality} \cite{}. Simply put, definitional
equality is equality through definition, which requires no proof. For example,
we know that for any two natural numbers {\tt m} and {\tt n}, {\tt m `plus` n}
is equal to the sum of {\tt m} and {\tt n} by definition of {\tt `plus`}.

@@

Now, let's prove the symmetry of identity types using {\tt J}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J ? ?
\end{code}
\end{minipage}
\end{center}
Next, we must fill in both instances of {\tt ?} in the above proof. Similar to
previous proofs, the first of these is our motive and is simply a restatement of
our type statement. In these forms of proofs, we commonly annotate parameterized
identity types with their given equality (viz. {\tt x=y}):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J (λ {x y} x=y → y ≡ x) ?
\end{code}
\end{minipage}
\end{center}
Now, the last {\tt ?} must be filled in with a proof that our motive holds in
the case that {\tt y} is {\tt x}, that is, we must prove that
{\tt (x : A) → x ≡ x}, which is the type of {\tt refl}. Thus, our completed
proof is the following:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J (λ {x y} x=y → y ≡ x) refl
\end{code}
\end{minipage}
\end{center}
To reiterate, this is an example of propositional equality. If this were a
definitional or judgmental equality, we would not need to prove symmetry, since
these two forms of equality are automatically reflexive, symmetric, associative,
and transitive. This is the primary and most important difference between other
forms of equality and propositional equality.

@@

\subsection*{Natural Numbers}
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data ℕ : Set where
  0 : ℕ
  suc : ℕ → ℕ

@@

\subsection*{List/Orderings}
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data List (A : Set) : Set where
  Nil  : List A
  _::_ : A → List A → List A

@@

\subsection*{Extension to Homotopy Type Theory (maybe)}

>>>>

\section{Introduction}
We present an introduction to dependent types and (Martin-L\"of) type theory.
At its core, type theory seeks to serve as a possible foundation of mathematics,
one that is based on a different logical foundation. This proof-relevant
form of mathematics is the basis in which types can be used to model logical
propositions and is the core idea of type theory. As such, not only does type
theory have a viable application in mathematics but also a growing interest
in the field of computer science where types find themselves in great use in
many popular programming languages. In both fields, type theory provides a
richer world of types and expressibiilty.

The study of type theory goes hand in hand with the study of dependent types,
which is why we first take the time to familiarize ourselves with them
(Section 2). We then continue with the fundamental concepts of type theory of
Per Martin-L\"of \cite{}. For this, we recommend the reader have a running proof
assistant, such as Agda. In our examples, we refer to Agda syntax rules, and
using a proof assistant should help develop the necessary thought processes to
successfully do type theoretic proofs.

@@

What follows is a brief description of each form of dependency and the
corresponding \(\lambda\)-calculus that is associated with it. With this, we
introduce the (fully-dependent) system of the \textit{Calculus of Constructions}
(\(\lambda C\)) \cite{}. Here, the word \textit{fully} signifies that \(\lambda C\)
includes all four forms of dependency mentioned above.

@@

The STLC has limited expressibilty but has the benefit of no longer having any
infinite calculations \cite{} and is thus the appropriate foundation for
deriving a type theoretic system. This is because the STLC lends itself quite
seamlessly as component of larger, more expressive systems. This flexibility is
what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
with more expressive types through simple extentions of its grammar of simple
types.

The key difference between the STLC and \(\lambda C\) are the places in which
terms and types are allowed to be. In the STLC, terms and types are entirely
seperate entities in that terms only exist on {\em term level} and types
on {\em type level}. We return to the idea of {\em levels} later. Thus, as we
mentioned earlier, this puts a constraint on the system to only allow
\(\lambda\)-abstraction and function application to happen at term level
(viz. definition of simple type). For example, the \textit{constant function}
over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC is defined:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
const : ℕ → 𝔹 → ℕ
const = λ n b → n
\end{code}
\end{minipage}
\end{center}
Given that {\tt const} is of type {\tt ℕ → 𝔹 → ℕ}, the term that inhabits this
type is {\tt λ n b → n}, where {\tt n} and {\tt b} are typed {\tt ℕ} and {\tt 𝔹},
respectively. Here, we are also limited by the fact that this constant function can
only be applied to natural numbers and booleans. It is not the most general
(i.e., applicable to any two terms) way to define such a function. Thus, in this
example, we can see that terms and types are treated differently and are only
related in terms of {\em inhabitance}. This relation is the reason for the
limited expressivity of the STLC and is also the method in which the STLC can be
extended into a more expressive system.

@@

Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
\(\lambda2\) as an extension of the STLC. This system is adds {\em polymorphic}
terms to the STLC (Dependency (2)). For example, let's take the definition of
{\tt const} above and change it to define {\tt polyConst}:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
polyConst : (A B : Set) → A → B → A
polyConst A B = λ a b → a
\end{code}
\end{minipage}
\end{center}
Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
are types. These two types (viz. {\tt A} and {\tt B}) are the corresponding
types of the third and fourth arguments of {\tt polyConst}. To clarify, we can
instantiate {\tt polyConst} with \(\mathbb{N}\) and \(\mathbb{B}\) resulting
in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us:
$${\tt const} = {\tt polyConst(ℕ,𝔹)}.$$
Thus, adding polymorphism to the STLC has the effect of extending the notion
of type and term. With \(\lambda2\), we are now allowed to pass types as arguments
at term level (viz. we pass the actual types \(\mathbb{N}\) and \(\mathbb{B}\)).
It is, however, necessary to constrain this expressivity with a more expressive
type (e.g. {\tt (x : Set) → ...}). Types that appear in this fashion are
called \(\Pi\)-types. Extending the grammar for simple types with \(\Pi\)-types
results in the grammar for types ({\tt t}) in \(\lambda2\):

@@

\(\Pi\)-types are actually significantly more expressive than the ones in
\(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). Before we can used
\(\Pi\)-types to their full potential, we must first introduce other forms of
type and term dependency.

@@

\begin{center}
\begin{minipage}{0.65\textwidth}
\begin{code}
Prim || t → t || ∀ (x : Set) → t || Set
\end{code}
\end{minipage}
\end{center}
The curious effect of doing this allows us to construct expressions of type
\mbox{{\tt t → t}}, where either {\tt t} can have subexpressions containing {\tt Set}.
Let's take the familiar definition of the list data structure:

@@

We discuss {\tt List} more in depth in a later section. For now, we simply ask
{\em what is the type of {\tt List}?} To answer this question, we can simply
look at the first line of the definition above (viz. {\tt List (A : Set) : Set}).
From this line, we can see that {\tt List} is a type parameterized over any type
{\tt A}. Another way to think about this statement is to think of {\tt List} as
a form of polymorphic function of type {\tt Set → Set}, as is similar with other
{\em parameterized types}. Unlike like {\tt const} or {\tt polyConst}, this
function can only be applied at type level (thus the name {\em type constructor}).
This is to insure that we cannot write type expressions that fail to instantiate
{\tt List}. For example, the type expression {\tt List → A} is incorrect since
{\tt List} cannot appear in a type expression without first being instantiated.
To correct this, we can write this type expression as {\tt List A → A}.

@@

To review, if we compile all the previously mentioned systems above, we have
the system \(\lambda\omega\) (System F\(\omega\)). Our final addition, the
system of \(\lambda P\), allows us to construct {\em type families} which are
themselves types and are meant to directly correspond to propositions in
higher-order logic.

@@

\begin{center}
\begin{minipage}{0.7\textwidth}
\begin{code}
data _isEven : ℕ → Set where
  ZEven : 0 isEven
  NEven : ∀ n → n isEven → (Suc (Suc n)) isEven
\end{code}
\end{minipage}
\end{center}
Here, we've defined two ways in which {\tt n isEven} for all natural numbers,
{\tt n}, can be proven. Returning to the idea of proof-relevant mathematics,
we do this by use of its two type constructors {\tt ZEven} and {\tt NEven},
corresponding to the two constructors of \(\mathbb{N}\). {\tt ZEven} is the
proof that 0 is an even number. {\tt NEven} then defines the property that if
a given natural number, {\tt n}, is even (viz. {\tt n isEven}), then {\tt n + 2}
is even (viz. {\tt (Suc (Suc n)) isEven}). To clarify how this type can be used,
let's prove that 4 is even. We start by providing the appropriate type definition
and also providing a name for our proof (viz. {\tt 4isEven}):
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
\end{code}
\end{minipage}
\end{center}
From the definition of {\tt isEven}, we know that to prove 4 is even, we must
know that 2 is even. Consequently, to prove 2 is even, we must know that 0 is
even. Fortunately, we have a proof that 0 is even: {\tt ZEven}. Thus, the proof
that 2 is even is {\tt NEven 0 ZEven}, and finally the proof that 4 is even is:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
4isEven = NEven 2 (NEven 0 ZEven)
\end{code}
\end{minipage}
\end{center}

@@

While it might seem intuitive and more effective to instead try to prove that 5
is {\em not} even, this, however, also does not work. This is due to the
treatment of negation in Intuitionistic Logic. To prove that 5 is not even is
to prove that \(\neg (5\) {\tt isEven}), which is equivalent to constructing a
function of type:
\begin{center}
\begin{minipage}{0.3\textwidth}
\begin{code}
5 isEven → ⊥
\end{code}
\end{minipage}
\end{center}
That is, given a proof that {\tt 5 isEven}, construct \(\perp\), the empty type.
This is not possible since we cannot ever construct a proof that 5 is even. We
discuss the empty type later in Section 3.

@@

\section{Martin-L\"of Type Theory}
%% Curry-Howard?
In this section, we introduce the fundamental concepts of Martin-L\"of Type
Theory \cite{}. We discuss recursion, induction, and as well several primitive type
encodings (from propositional logic).

@@

\begin{minipage}{0.7\textwidth}
\begin{small}
The recursion and induction principles for some type {\tt T}.
\end{small}
\end{minipage}
\end{center}
The first argument given to these principles is sometimes called the {\em motive}.
This simply means that {\tt C} is the type from which we will return an element
from. For recursion principles, this means the given output of the function we
are defining, and for induction principles, this can be read as the proposition
we are trying to prove. The terms that follow (viz. {\tt C₀}) correspond to the
constructor(s) of the given type, which is how these principles guarantee
totality. Finally, both principles take a term of the type that are acting on
(viz. {\tt T} and {\tt (t : T)}). The only difference is that the induction
principle introduces a type family (e.g. similar to {\tt isEven}), while a
recursion principle does not.

@@

prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
prop₀ ¬¬¬a a = ?
\end{code}
\end{minipage}
\end{center}
Now, we solve for {\tt ?}. Looking at the type expression of {\tt prop₀},
we know we have an expression of type \(\neg\neg\neg A\) and also an
expression of type \(A\). To clarify, we can rewrite the type expression
after desugaring \(\neg A\):
$$\neg\neg\neg A\rightarrow A\rightarrow\perp$$
Now, to construct an element of type \(\neg A\) from an element of type
\(\neg\neg\neg A\), we can use the definition of \(\neg\):
$$(\neg\neg A\rightarrow\perp)\rightarrow A\rightarrow\perp$$
$$((\neg A\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
$$(((A\rightarrow\perp)\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
Thus, we know we must apply our expression \(\neg\neg\neg A\) to a function
of type \(\neg\neg A\) and so on, such that:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
prop₀ ¬¬¬a a = ¬¬¬a (λ ¬a → ¬a a)
\end{code}
\end{minipage}
\end{center}
Or equivalently:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
prop₀ ¬¬¬a a = rec⊥ ⊥ (¬¬¬a (λ ¬a → ¬a a))
\end{code}
\end{minipage}
\end{center}
For clarity, we annotate terms with their respective type (i.e., {\tt ¬¬¬a} is
of type {\tt ¬¬¬A}). We do this quite often when writing proofs.

Next, we introduce the type \(\top\), which contains exactly one element,
{\tt *}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data ⊤ : Set where
  * : ⊤

@@

ind⊤ : (C : ⊤ → Set) → C * → (x : ⊤) → C x
ind⊤ C c * = c
\end{code}
\end{minipage}
\end{center}
Similar to \(\perp\), \(\top\) is simple and is rather uninteresting by itself.
There are, however, some trivial yet interesting proofs that involve the two.
For the first of these, we also introduce a special kind of proof: {\it uniqueness
of identity} (UIP) \cite{}. In this proof, we set out to determine and classify
the identity of every element within a given type. For example, for \(\top\), we
should be able to prove:
$$\forall (x : \top) \rightarrow x ≡ *$$
That is, for every element of type \(\top\), that element is equivalent to {\tt *}.
This second expression, {\it x ≡} {\tt *}, is an example of an {\it identity type},
which we describe later on (Section 4). For now, we simply say that the
sole constructor for the identity type is {\tt (refl x)}, which is a proof that
{\it x ≡ x}, for any term {\tt x}.

Now, to prove the above proposition, we can use the induction principle for
\(\top\), since the proposition we are trying to prove concerns all elements
of \(\top\):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
uip⊤ : (x : ⊤) → x ≡ *
uip⊤ = ind⊤ (λ top → top ≡ *) (refl *)
\end{code}
\end{minipage}
\end{center}
Here, we have taken our motive to be the expression {\tt (λ top → top ≡ *)},
which essentially repeats the information already given by our type expression
(viz. {\tt (x : ⊤) → x ≡ *}). The second element, {\tt (refl *)}, is a proof
that our motive holds in the case that {\tt top} is {\tt *} and is of type
{\tt * ≡ *}.

@@

ind+ : {A B : Set} (C : A + B → Set) →
       (∀ (x : A) → C (inl x)) →
       (∀ (x : B) → C (inr x)) →
       (x : A + B) → C x
ind+ C f _ (inl a) = f a
ind+ C _ f (inr b) = f b
\end{code}
\end{minipage}
\end{center}
It might come as surprise as to why the symbol for addition is generally used
for the disjunction type. This is because another name for + is a {\it union type}.
This type is parameterized over two other types, {\tt A} and {\tt B}, and
creates a union between them in that every expression of type {\tt A + B} is
either {\tt (inl a)} or {\tt (inr b)}, where {\tt a} and {\tt b} are typed
{\tt A} and {\tt B}, respectively.

@@

We can now write a simple proof that shows the union of \(\top\) and \(\top\)
is equivalent to the Boolean type, that is {\tt (⊤ + ⊤) ↔ 𝔹}.
Alternatively, since \(\top\) is occasionally written as 1 and {\tt 𝔹} as 2, we
name our proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
1+1=2 : (⊤ + ⊤) ↔ 𝔹
1+1=2 = ?
\end{code}
\end{minipage}
\end{center}
To do this, we must associate every possible term of type {\tt ⊤ + ⊤} to a
given Boolean. This works out quite simply since the type {\tt ⊤ + ⊤} only has
two possible terms, {\tt (inl *)} and {\tt (inr *)}, which we can associate with
the two terms, {\tt True} and {\tt False}, in the type {\tt 𝔹} and vice-versa.
Thus, we can write the proof as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
1+1=2 = <> (f , g)
  where f = rec+ 𝔹 (λ inl* → True) (λ inr* → False)
        g = rec𝔹 (⊤ + ⊤) (inl *) (inr *)
\end{code}
\end{minipage}
\end{center}
For our purposes, we use the constructor {\tt <>} for constructing a term of
the bijection type. For simplicity, we can define a bijection as a conjunction
of two logical implications:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data _↔_ (A B : Set) : Set where
  <>_ : (A → B) × (B → A) → (A ↔ B)
\end{code}
\end{minipage}
\end{center}
To clarify, this form of equivalence is called ``judmental equality'' (i.e.,
equality by logical bijection) \cite{}. We go over the three different forms of
equivalence in the following section.

@@

\subsection*{Equality and Identity Types}
%% talk about definitional equality, judgmental equality, prop equality
The first form of equality is much simpler than the others. This form of
equality is called {\it definitional equality} \cite{}. Simply put, definitional
equality is equality through definition, which requires no proof. For example,
we know that for the natural numbers, {\tt 1 `plus` 1} is equal to {\tt 2} by
the definition of {\tt plus}.

@@

Now, let's prove the symmetry of identity types using {\tt J}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J ? ?
\end{code}
\end{minipage}
\end{center}
Next, we must fill in both instances of {\tt ?} in the above proof. Similar to
previous proofs, the first of these is our motive and is simply a restatement of
our type statement. As we have done earlier, we annotate parameterized identity
types with their given equality (viz. {\tt x=y}):
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J (λ {x y} x=y → y ≡ x) ?
\end{code}
\end{minipage}
\end{center}
Now, the last {\tt ?} must be filled in with a proof that our motive holds in
the case that {\tt y} is {\tt x}, that is, we must prove that
{\tt (x : A) → x ≡ x}, which is the type of {\tt refl}. Thus, our completed
proof is the following:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
sym = J (λ {x y} x=y → y ≡ x) refl
\end{code}
\end{minipage}
\end{center}
To reiterate, this is an example of propositional equality. If this were a
definitional or judgmental equality, we would not need to prove symmetry, since
these two forms of equality are automatically reflexive, symmetric, associative,
and transitive. This is the primary and most important difference between other
forms of equality and propositional equality.

@@

\subsection*{Natural Numbers}
Natural numbers have two constructors: {\tt 0} and {\tt suc}. The latter of
these is a function that simply adds 1 to any natural number. Natural numbers
are the simplest example of an infinite type (i.e., it encodes every positive
whole number, {\tt n}, such that {\tt n} greater than or equal to {\tt 0}).

\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data ℕ : Set where
  0 : ℕ
  suc : ℕ → ℕ

@@

\subsection*{List/Orderings}
Lists are similar to natural numbers in that their type definition also has two
constructors: {\tt Nil}, which represents the empty list, and a function that
adds an element to the front of a given list, commonly read as {\em cons}.

\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
data List (A : Set) : Set where
  Nil  : List A
  _::_ : A → List A → List A

@@

\subsection*{Extension to Homotopy Type Theory (maybe)}
\subsection*{Conclusion and Acknowledgments}
